### 2018-06-26 latex/results.tex
revised results section

NAD can be synthesized using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA) (Overview see Fig 1.). Nam and NA are together known as vitamin B3 or niacin. Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Yoshino2018}. Du to the hight turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursor of human NAD-biosynthesis.

@@

First we addressed the question of the predominant coexitence of NamPRT and NNMT in vertebrates that coincides with an increase in the number of NAD-consuming enzyme families. We thus used our dynamic model to simulate steady state concentrations and NAD-consumption rates simulating NAD biosynthesis via  NamPRT in the  presence or absence of NNMT.  We see that the presence of NNMT enables higher NAD consumption rates  (Figure3A) while, as exptected, reducing the steady state concentration of NAD (Figure 3B). The decline in NAD-concentration can be compensated by a higher expression of NamPRT further increasing NAD consumption flux.

These findings can be explained when looking in more detail into the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirt1p.  The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, explaing why the presence of NNMT enables higher NAD consumption fluxes. At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD-concetration.

As kinetic parameters of NamPRT are only available for the human enzyme \cite{Burgos2008} , we went on to theoretically analysed the potential effect of the NamPRT affinity (Km) on NAD steady state concentration and NAD consumption flux, leaving all other kinetic parameters constant. In the absence of NNMT (Figure 3C-D) a change in Km has very little effect on steady state NAD concentration and NAD consumption flux.  In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing Km values (Figure 5E-F).

It is intersting to note that, without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant. Even though, this is of course an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead. However, there seems to be an apparent trade off between maintainable NAD concentration and consumption flux.

When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at a low affinity (Km of 1 mikroM, which is in the range of the Km of NADA for Nam, NAD consumption flux is only higher with NNMT at low cell division rates. This effect becomes stronger if we further lower the affinity (not shown) (Figure 3G-H).

The pathway dynamics are of course not solely dependent on one enzyme. Thus, what is the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate? In Figure 4A we see that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes. Thus, a further increase of the affinity of NamPRT for Nam would not provide any advantage.

@@

As our simulations suggest that NNMT might have excerted an evolitionary pressure on the development of NamPRT. We performed multiple sequence analysis to see if we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT. The multiple sequence alignment of selected sequences is shown in Figure 5A. A more comprehensive multiple sequence alignment containing a larger number of species can be found in Supplementary Figure 1.  We recognized that most Deuterostomia that have only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme. This insert overlaps with a predicted weak nuclear localisation signal, that is lost when removing this sequence stretch. The inserted sequence corresponce to a structurallly unresolved loop structure in all avaibable crystal structure of human NamPR (e.g. \cite{Wang2006} structure visualisation Figure 5B). The loop is connected to one of the beta-sheets involved in substrate binding.

From these observations we derived two hypotheses regarding the potential function of the loop for NamPRT function. Firstly, we analysed whether the deletion of the amino acids 42 to 51 has an effect on the localisation of the human enzyme. But both wildtype and mutant protein show a mixed localisation in the cytosol and nucleus (Figure 6C). We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the sequence deletion.
The second hypothesis was that the sequence insertion influences NamPRT affinity and activity, which is also what our mathematical simulations would predict. We thus recombinantly expressed  both wildttype and mutant  NamPRT  in E. coli.  Upon purifictaion we meassured the activity using NMR detecttion of NMN produced in the presence and absence of ATP. It appears that the enzymatic activity of the mutant enzyme is much lower than that of the wildtype enzyme (Figure 6D). We furthermore see that in contrast to NamPRT wildtype, the mutant does not show an increased activity in the presence of ATP (Figure 6E).

@@

But why is NADA lost in vertebrates while NADA and NamPRT are co-existing existing in bacteria? To analyse this question we built a two compartment model that has a shared Nam source (details see Materials an Methods and Supplementary Material). One compartment contains NADA while the other contains either NamPRT alone or together with NNMT.  Without NNMT the compartment containing NADA has slightly lower NAD consumption rates (Figure 7A), but is able to maintain much higher steady state NAD concentrations at low cell division rates  (Figure 7B). At high cell division rates, however, steady state concentration in both compartments are similar. The latter might explain why in bacteria that have relatively high growth rates both systems coexists. 
If in contrast, NNMT is present in the NamPRT compartment, this compartment outcompetes the NADA containing compartment, both with respect to NAD-consumption rates and steady state NAD concentrations (Figure 7 C and D). This effect is dependent on a high affinity of NamPRT, as at low affinity the NADA containing system again maintains higher NAD-concentrations, the NAD consumption flux still being higher  in the NamPRT and NNMT containing compartment. Taken together this might explain why upon the development of a sufficiently high affinity of NamPRT for its substrate, which seems to have been induced by the appearance of NNMT, NADA is lost.

>>>>

NAD can be synthesized using several routes from three main precursors: tryptophan, nicotinamide (Nam) and nicotinic acid (NA) (Overview see Figure 1.). Nam and NA are together known as vitamin B3 or niacin. Alternatively, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead \cite{Yoshino2018}. Du to the hight turnover of NAD observed and the fact that only 1\% of the tryptophan taken up with our diet is converted into NAD, vitamin B3 (Nam and NA) and to a lower extend NR are the major precursor of human NAD-biosynthesis.

@@

First we addressed the question of the predominant coexitence of NamPRT and NNMT in vertebrates that coincides with an increase in the number of NAD-consuming enzyme families. We thus used our dynamic model to simulate steady state concentrations and NAD-consumption rates simulating NAD biosynthesis via  NamPRT in the  presence or absence of NNMT.  We see that the presence of NNMT enables higher NAD consumption rates  (Figure3A) while, as exptected, reducing the steady state concentration of NAD (Figure 3B). The decline in NAD-concentration can be compensated by a higher expression of NamPRT further increasing NAD consumption flux (Figure 3A and B dashed lines).

These findings can be explained when looking in more detail at the kinetic parameters of NamPRT and NAD-consuming enzymes such as Sirt1p.  The increase of NAD consumption flux is caused by the fact that most NAD-consuming enzymes are inhibited by their product Nam, explaing why the presence of NNMT enables higher NAD consumption fluxes. At the same time, the high substrate affinity of NamPRT maintains a sufficiently high NAD-concetration.

As kinetic parameters of NamPRT are only available for the human enzyme \cite{Burgos2008}, we went on to theoretically analysed the potential effect of the NamPRT affinity ($K_M$) on NAD steady state concentration and NAD consumption flux. In the absence of NNMT (Figure 4A-B) a change in $K_M$ has very little effect on steady state NAD concentration and NAD consumption flux.  In the presence of NNMT, however, NAD consumption flux and NAD concentration increases with decreasing $K_M$ values (Figure 4C-D).

It is intersting to note that, without NNMT, NAD concentration and consumption flux are both considerably affected by cell division rates, at least if the enzyme expression is kept constant. This is of course an artificial scenario, as one would assume organisms to regulate enzyme expression to achieve similar levels of metabolite concentrations instead. However, there seems to be an apparent trade off between maintainable NAD concentration and consumption flux in the absecne of NNMT. In contrast,  NAD consumption rates increase with NAD concentration in the presence of NNMT and both are relatively independent of cell devision rates.

When we compare NAD consumption and NAD concentration with and without NNMT with two different substrate affinities of NamPRT, we see that at low affinity ($K_M$ of 1 $\mu$M, which is in the range of the $K_M$  of NADA for Nam), NAD consumption flux is only higher with NNMT at low cell division rates. This effect becomes stronger with decreasing affinity (Figure 4E-F).

The pathway dynamics are of course not solely dependent on one enzyme. Thus, what is the impact of the substrate affinity of NNMT that is competing with NamPRT for the same substrate? In Figure 5 we see that the substrate affinity values found in the human enzymes (indicated by black asterisks) are actually optimal with respect to both achievable steady state NAD concentration and consumption fluxes. Thus, a further increase of the affinity of NamPRT for Nam would not provide any advantage.

@@

As our simulations suggest that NNMT might have excerted an evolutionary pressure on the development of NamPRT. We performed multiple sequence analysis to see if we can find sequence variations in the protein sequence of NamPRT that indicate evolutionary changes of NamPRT upon the appearance of NNMT. The multiple sequence alignment of selected sequences is shown in Figure 5A and a more comprehensive multiple sequence alignment containing a larger number of species can be found in Supplementary Figure 1.  We recognized that most Deuterostomia that have only NamPRT and NNMT (indicated by the blue circle) have an insert of ten amino acids corresponding to positions 43 to 52 of the human enzyme. This insert overlaps with a predicted weak nuclear localisation signal, that is lost when removing this sequence stretch. The inserted sequence corresponce to a structurallly unresolved loop in all avaibable crystal structure of human NamPR (e.g. \cite{Wang2006} structure visualisation Figure 5B). The loop is connected to one of the beta-sheets involved in substrate binding.

From these observations we derived two possible hypotheses regarding the role of the loop for NamPRT function. Firstly, we analysed whether the deletion of the amino acids 42 to 51 has an effect on the localisation of the human enzyme. We thus performed immunoflourescence imaging with FLAG-tagged proteins, but both wildtype and mutant protein showed a mixed cytosolic nuclear localisation (Figure 6C). We therefore conclude that the partial nuclear localisation of NamPRT is not affected by the sequence deletion.
The second hypothesis was that the sequence insertion influences NamPRT affinity, which is what our mathematical simulations would predict. We thus recombinantly expressed  both wildttype and mutant  NamPRT  in E. coli.  Upon purifictaion we meassured the activity using NMR detecttion of NMN produced in the presence and absence of ATP. It appears that the enzymatic activity of the mutant enzyme is much lower than that of the wildtype enzyme (Figure 6D) and that an increase of the substrate concentration has a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity.  We furthermore showed  that in contrast to NamPRT wildtype, the mutant does not exhibit an increased activity in the presence of ATP (Figure 6E).

@@

The remaining question is now, why NADA is lost in vertebrates, while NADA and NamPRT are co-existing in bacteria? To analyse this question we built a two compartment model that has a shared Nam source (details see Materials an Methods and Supplementary Material). One compartment contains NADA while the other contains either NamPRT alone at equal expression levels, or together with NNMT.  Without NNMT the compartment containing NADA has slightly lower NAD consumption rates (Figure 7A), but is able to maintain much higher steady state NAD concentrations at low cell division rates  (Figure 7B). At high cell division rates, steady state concentrations in both compartments are similar. The latter might explain why in bacteria that have relatively high growth rates both systems co-exist. 
In the presence of NNMT the NamPRT compartment outcompetes the NADA containing compartment, both with respect to NAD consumption rates and steady state NAD concentrations (Figure 7 C and D). This effect is dependent on a high affinity of NamPRT, as at low affinity (high $K_M$ ) the NADA containing system is again able to maintain higher NAD-concentrations. The NAD consumption flux is still  higher  in the NamPRT and NNMT containing compartment. Taken together this might explain why upon the development of a sufficiently high affinity of NamPRT for its substrate, which seems to have been induced by the appearance of NNMT, NADA is lost.
